\documentclass[sigconf]{acmart}

\begin{document}

%\markboth{V. F. Pamplona et al.}{Photorealistic Models for Pupil Light Reflex and Iridal Pattern Deformation}

\title{Towards a Methodology and Framework for AI Sustainability Metrics} % title

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
% \author{Ben Trovato}
% \authornote{Both authors contributed equally to this research.}
% \email{trovato@corporation.com}
% \orcid{1234-5678-9012}
% \author{G.K.M. Tobin}
% \authornotemark[1]
% \email{webmaster@marysville-ohio.com}
% \affiliation{%
%   \institution{Institute for Clarity in Documentation}
%   \streetaddress{P.O. Box 1212}
%   \city{Dublin}
%   \state{Ohio}
%   \country{USA}
%   \postcode{43017-6221}
% }

% \author{Lars Th{\o}rv{\"a}ld}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{larst@affiliation.org}

% \author{Valerie B\'eranger}
% \affiliation{%
%   \institution{Inria Paris-Rocquencourt}
%   \city{Rocquencourt}
%   \country{France}
% }

% \author{Aparna Patel}
% \affiliation{%
%  \institution{Rajiv Gandhi University}
%  \streetaddress{Rono-Hills}
%  \city{Doimukh}
%  \state{Arunachal Pradesh}
%  \country{India}}

% \author{Huifen Chan}
% \affiliation{%
%   \institution{Tsinghua University}
%   \streetaddress{30 Shuangqing Rd}
%   \city{Haidian Qu}
%   \state{Beijing Shi}
%   \country{China}}

% \author{Charles Palmer}
% \affiliation{%
%   \institution{Palmer Research Laboratories}
%   \streetaddress{8600 Datapoint Drive}
%   \city{San Antonio}
%   \state{Texas}
%   \country{USA}
%   \postcode{78229}}
% \email{cpalmer@prl.com}

% \author{John Smith}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{jsmith@affiliation.org}

% \author{Julius P. Kumquat}
% \affiliation{%
%   \institution{The Kumquat Consortium}
%   \city{New York}
%   \country{USA}}
% \email{jpkumquat@consortium.net}

\begin{abstract}
Recently, we are witnessing truly groundbreaking achievements using AI models, 
such as the much talked about generative large language models, the broader area of foundation models, and the wide range of applications 
with a tremendous potential to accelerate scientific discovery, and enhance productivity. 
AI models and their use are growing at a super-linear pace leading to staggering environmental footprint implications. Inference jobs are measured by the
trillions, and model parameters by the billions. This scaling up comes with a tremendous 
environmental cost, associated with every aspect of models' life cycle: data preparation, pre-training, and post deployment re-training, inference, and, the embodied emission of the systems used to support the execution. There is an urgent need for the community to come together and conduct a meaningful conversation about the environmental cost of AI. To do that, we ought to develop an agreed upon set of metrics, methodology, and framework to quantify AI's sustainability cost in a holistic and complete fashion. 
%Moreover, 
%recent advances brought to the front the concept of foundation models, promoting model re-use through a process of distillation and fine tuning, along with a debate around the most cost effective strategy: downstream distillation of large broad models vs. train-from-scratch of smaller focused models. A unified sustainability metric can enable an informed  debate about this question.  
In this paper, we propose unified AI Sustainability metrics that can help foster a sustainability mind-set and enable analysis, and strategy setting. To do that, we build on and extend the data center sustainability metrics, defined in ~\cite{gandhi2022metrics}, by introducing (for the first time to our knowledge) the concept of {\em embodied product emission (EPC)} to capture the creation cost of software assets, such as software platforms, models, and data-sets. We then use this new concept to expand the job sustainability cost metrics ($JCS$ and $ASC$) offered in~\cite{gandhi2022metrics} to factor in 
the context of the execution of jobs, such as a platform as-a-service, or a model serving inference jobs. The result is applicable to any data center job, not just for AI, and contributes towards accuracy and completeness. 
We then show how to apply this approach to AI, with a particular
focus on the entire life cycle, including all phases of the life cycle, as well as the provenance of models, where one model is used (distilled) to create another one. We demonstrate how the metric can be used to inform a more meaningful debate about AI strategies and cost.  
\end{abstract}

\keywords{sustainable computing, green AI, sustainable AI, foundation models}
\maketitle

\input{background}

\input{metric.tex}

\input{opportunities}

\appendix

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{acmtog-sample-bibfile}

\end{document}

